---
title: "Battleship Game"
author: "KN Mnisi"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
_____________________________________________________________________________________________________
Packages
```{r}
library(sampling)
```
_____________________________________________________________________________________________________
Setting up game board dimensions
```{r}
board_size <- 10
num_ships <- 5
ship_lengths <- c(5, 4, 3, 3, 2)
num_games <- 20
```
_____________________________________________________________________________________________________
Function to Create empty game boards
```{r}
#for (i in 1:num_games) {
createBoard <- function(){
col_names <- c("1","2","3","4","5","6","7","8","9","10")
row_names <- c("A","B","C","D","E","F","G","H","I","J")
board <- matrix(rep(0, board_size^2), nrow = board_size)
rownames(board) <- row_names
colnames(board) <- col_names
return(board)
}
```
_____________________________________________________________________________________________________
Function to randomly place ships on a board without overlapping
```{r}
place_ships <- function(board, ship_lengths) {
  for (ship_length in ship_lengths) {
    ship_placed <- FALSE
    while (!ship_placed) {
      row_start <- sample(1:nrow(board), 1)
      col_start <- sample(1:ncol(board), 1)
      direction <- sample(c("horizontal", "vertical"), 1)
      
      # Checking that ship doesn't overlap with edges of board
      if (direction == "horizontal" &&
          col_start + ship_length - 1 <= ncol(board) &&
          all(board[row_start, col_start:(col_start + ship_length - 1)] == 0)) {
        board[row_start, col_start:(col_start + ship_length - 1)] <- 1
        ship_placed <- TRUE
      } else if (direction == "vertical" &&
                 row_start + ship_length - 1 <= nrow(board) &&
                 all(board[row_start:(row_start + ship_length - 1), col_start] == 0)) {
        board[row_start:(row_start + ship_length - 1), col_start] <- 1
        ship_placed <- TRUE
      }
      if (direction == "horizontal" &&
          col_start + ship_length - 1 <= ncol(board) &&
          all(board[row_start, col_start:(col_start + ship_length - 1)] == 0)) {
        board[row_start, col_start:(col_start + ship_length - 1)] <- 1
        ship_placed <- TRUE
      } else if (direction == "vertical" &&
                 row_start + ship_length - 1 <= nrow(board) &&
                 all(board[row_start:(row_start + ship_length - 1), col_start] == 0)) {
        board[row_start:(row_start + ship_length - 1), col_start] <- 1
        ship_placed <- TRUE
      }
    }
  }
  return(board)
}
```
_____________________________________________________________________________________________________
Function to display a board
```{r}
display_board <- function(board) {
  board_display <- board
  board_display[board_display == 1] <- 0 # Hide ships
  board_display[board_display == 2] <- "X" # Show hits
  board_display[board_display == 3] <- "O" # Show misses
  print(board_display)
}
```
_____________________________________________________________________________________________________
Function to check if a shot hits a ship
```{r}
check_shot <- function(opponentBoard, row, col) {
  if (opponentBoard[row, col] == 1) {
    cat("Hit!!!\n")
    opponentBoard[row, col] <- 2 # Mark hit
  } else {
    cat("Missed!!!\n")
    opponentBoard[row, col] <- 3 # Mark miss
  }
  return(opponentBoard)
}
```
_____________________________________________________________________________________________________
Function to check if a player has won
```{r}
checkWin <- function(opponentBoard) {
  if (1 %in% opponentBoard) {
    return(FALSE)
  } else {
    return(TRUE)
  }
}
```
_____________________________________________________________________________________________________
Function to play the game
```{r}
playGame <- function() {
  # Initialize the number of wins
  num_wins_player1 <- 0
  num_wins_player2 <- 0
  shots_taken_in_round <- 0

  # Initialize sets to track shots taken by each player
  shots_taken_by_player1 <- list()
  shots_taken_by_player2 <- list()

  # Display the initial game boards
  cat("Player 1 Board:\n")
  display_board(my_board)
  cat("Player 2 Board:\n")
  display_board(opp_board)

  # Start the game
  currentPlayer <- 1
  winner <- NULL
  repeat {
    if (currentPlayer == 1) {
      cat("Player 1's turn:\n")
      targetBoard <- opp_board
      shots_taken <- shots_taken_by_player1
    } else {
      cat("Player 2's turn:\n")
      targetBoard <- my_board
      shots_taken <- shots_taken_by_player2
    }

    # Take a random shot
    repeat {
      row <- sample(LETTERS[1:10], 1)
      col <- sample(1:10, 1)
      shot <- paste(row, col, sep = "")

      # Check if the shot has already been taken
      if (!(shot %in% shots_taken)) {
        shots_taken <- c(shots_taken, shot)
        break
      }
    }
    cat("Shot taken at: ", shot, "\n")
    shots_taken_in_round <- shots_taken_in_round + 1

    # Check if the shot hit or missed
    opp_shot <- check_shot(targetBoard, row, col)

    # Check if the current player has won
    if (checkWin(opp_shot)) {
      if (currentPlayer == 1) {
        cat("Player", currentPlayer, "WON!!!!\n")
        opponent <- 3 - currentPlayer
        cat("Player's", opponent, "Board\n")
        display_board(opp_shot)
        cat("Shots Taken by Player", currentPlayer, ":", round((shots_taken_in_round/2),0), "\n")
        num_wins_player1 <- num_wins_player1 + 1
        winner <- "Player 1"
      } else {
        cat("Player", currentPlayer, "WON!!!!\n")
        opponent <- 3 - currentPlayer
        cat("Player's", opponent, "Board\n")
        display_board(opp_shot)
        cat("Shots Taken by Player",currentPlayer ,":", round((shots_taken_in_round/2),0), "\n")
        num_wins_player2 <- num_wins_player2 + 1
        winner <- "Player 2"
      }
      break
    }

    # Check if the game is over
    if (!1 %in% opp_shot) {
      break
    }

    # Update the respective player's board
    if (currentPlayer == 1) {
      opp_board <- opp_shot
      shots_taken_by_player1 <- shots_taken
    } else {
      my_board <- opp_shot
      shots_taken_by_player2 <- shots_taken
    } 

    # Display the updated game board
    opponent <- 3 - currentPlayer
    cat("Player", opponent,"'s", "Board\n")
    display_board(opp_shot)
    # Switch to the next player
    currentPlayer <- 3 - currentPlayer
  }

    # Return number of wins for each player
  return(list(player1 = num_wins_player1, player2 = num_wins_player2, shots_in_round= shots_taken_in_round, winner= winner))
}
```
_____________________________________________________________________________________________________
Playing the game 'num_of_games' times and recording the results in a table
```{r}
# Initialize counters for total wins across all games
total_wins_player1 <- 0
total_wins_player2 <- 0

# Initialize data frame to store game results
game_results_random <- data.frame(
  `Game Number` = integer(),
  Winner = character(),
  `Shots Taken By Winner` = integer(),
  stringsAsFactors = FALSE
)

# Creating a loop to play the game many times
for (game in 1:num_games) {
  cat("\n===== Game", game, "=====\n")

# Create empty game boards for player 1 and player 2
cat("My Board","\n")
(my_board <- createBoard())
cat("Opponent's Board","\n")
(opp_board <- createBoard())

#Place ships on the game boards
cat("Player 1's Board","\n")
(my_board <- place_ships(my_board, ship_lengths))
cat("Player 2's Board","\n")
(opp_board <- place_ships(opp_board, ship_lengths))

# Play the game until either player wins
game_result <- playGame()
# Store the results in the data frame
game_results_random <- rbind(game_results_random, data.frame(
  `Game Number` = game,
  Winner = game_result$winner,
  `Shots Taken By Winner` = round((game_result$shots_in_round / 2), 0),
  stringsAsFactors = FALSE
))

# Accumulate wins for each player across all games
total_wins_player1 <- total_wins_player1 + game_result$player1
total_wins_player2 <- total_wins_player2 + game_result$player2
}
```
_____________________________________________________________________________________________________
Print the number of wins for the round of iterations and record the winning information for each game in a table
```{r}
# Print total wins for each player after all games
cat("\n===== Total Wins =====\n")
cat("Player 1 total wins:", total_wins_player1, "\n")
cat("Player 2 total wins:", total_wins_player2, "\n")
print(game_results_random)
```

